---
title: "Sample GEO RNA-seq Dataset Processing Code using DESeq2 (GSE124422)"
author: "Shayna Kay"
date: "07/04/2022"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

## Standard set-up of R Markdown environment

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls()) # Clear R environment's memory prior to beginning analysis 
```

## Install packages and load libraries
Required packages include: "htmltools", "DESeq2" from "Bioconductor", "ggplot2", and "dplyr".

```{r install, include=FALSE}
#install.packages("htmltools")
library(htmltools)

#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
# BiocManager::valid()
#BiocManager::install(c("biocLite"))
#BiocManager::install("DESeq2")
#biocLite("DESeq2")

library(DESeq2)
library(ggplot2)
library(dplyr)
```

## Read count data into RStudio and process data for analysis
Prepare the count data table as required for input into DESeq2 functions.

```{r data}
# file.choose()

# Read in .csv file containing RNA-seq count data
countData <- read.csv('C:\\Users\\shayn\\OneDrive\\Documents\\Thesis Documents\\DATA\\Bioinformatics\\GEO Files\\Select Datasets for Analysis 1\\3-GSE124422 - Copy_done\\GSE124422_htseq_counts_atra-ctrl.csv', header = TRUE, sep = ",") 

head(countData) # Print the first 6 lines of the count table to confirm data import was successful

countData <- countData %>% rename(ID = ï..ID) # Rename column containing gene symbols to "ID"
head(countData) # Print the first 6 lines of the count table to confirm processing was successful
```

## Generate a metadata table 
DESeq2 requires a metadata table containing information about the RNA-seq samples.

```{r metadata}
condition <- factor(c("X1_ATRA_counts","X2_ATRA_counts","X3_ATRA_counts","X1_untreated_counts","X2_untreated_counts","X3_untreated_counts")) # Condition column links metadata table to count table
treatment <- factor(c("ATRA", "ATRA", "ATRA", "control", "control", "control"))
GEO_ID <- factor(c("GSM3532958","GSM3532965", "GSM3532972", "GSM3532961", "GSM3532968", "GSM3532975"))
metadata <- DataFrame(condition, treatment, GEO_ID) # Create the dataframe containing above sample information
head(metadata) # Print the first 4 lines of the metadata table to confirm it was created successfully
```

## Construct DESEQDataSet Object to analyze differential gene expression
Export a CSV file containing results for further analysis of genes of interest.

```{r dds}
# The "DESeqDataSetfromMatrix" function stores intermediate calculations and results of the analysis of 
# differential expression based on a formula which specifies the experimental design
dds <- DESeqDataSetFromMatrix(countData=countData, 
                              colData=metadata, 
                              design=~treatment, tidy = TRUE) 

# Re-level experimental treatments, making control the reference treatment
dds$treatment <- relevel(dds$treatment, ref = "control") 
dds # View intermediate results of differential expression analysis

dds <- DESeq(dds) # Complete differential expression analysis and store output in "dds" variable
res <- results(dds) # Store the results of differential expression analysis in "res" variable
head(results(dds, tidy = TRUE)) # View the first 6 lines of the results table to confirm expected output

# Export the results of differential expression analysis and save on computer as a CSV file:
write.csv(res,"C:\\Users\\shayn\\OneDrive\\Documents\\Thesis Documents\\DATA\\Bioinformatics\\GEO Files\\Select Datasets for Analysis 1\\3-GSE124422 - Copy_done\\DESeqResults.csv") 
# Adjusted p-values<0.05 were extracted from this file to identify significantly differentially expressed
# sialyltransferase and neuraminidase genes 
# Log2(FoldChange) values were extracted from this file to create a heat map (Figure 10) of differential 
# sialyltransferase and neuraminidase gene expression

summary(res) # View summary of differential gene expression
```

# Create plots of normalized counts for sialyltransferase and neuraminidase genes of interest

```{r plotCounts}
# Use the "plotCounts" function to compare the normalized counts between treated and control groups

# Gemerate plots of normalized counts for ACTB and all sialyltransferases and neuraminidases with significant differences
par(mfrow=c(2,4)) # Set plot dimensions and layout
plotCounts(dds, gene="ACTB", intgroup="treatment")
plotCounts(dds, gene="ST3GAL1", intgroup="treatment")
plotCounts(dds, gene="ST3GAL3", intgroup="treatment")
plotCounts(dds, gene="ST3GAL4", intgroup="treatment")
plotCounts(dds, gene="ST6GAL1", intgroup="treatment")
plotCounts(dds, gene="ST6GALNAC6", intgroup="treatment")
plotCounts(dds, gene="NEU3", intgroup="treatment")
```

## Create a volcano plot
Visualize significantly differentially expressed genes in the entire dataset.

```{r volcano}
par(mfrow=c(1,1)) # Set plot dimensions and layout
with(res, plot(log2FoldChange, -log10(pvalue), pch=20, main="Volcano plot GSE124422", xlim=c(-3,3), ylim=c(0,200))) # Create plot with the given specifications

# Add coloured points: blue if padj<0.01, red if log2FC>1 and padj<0.05)
with(subset(res, padj<0.01), points(log2FoldChange, -log10(pvalue), pch=20, col="blue"))
with(subset(res, padj<0.01 & abs(log2FoldChange)>2), points(log2FoldChange, -log10(pvalue), pch=20, col="red"))
```

## Principal Component Analysis
Shows similarity of samples within the same group in transformed RNA-seq count data.

```{r PCA}
# Transform the raw count data
# The "vst" function will perform variance stabilizing transformation
vsdata <- vst(dds, blind=FALSE)
par(mfrow=c(1,1)) # Set plot dimensions and layout

# Using the "plotPCA" function from the "DESeq2" package to look at how samples group by treatment and day:
plotPCA(vsdata, intgroup="treatment") 
```